Validated Model Counting Data \\ Version of \today

### 2023-02-27 doc/mc-data.tex
Data update

\section*{Methodology}
\begin{enumerate}
   \item Retrieved 200 public model counting benchmarks from 2022 Model Counting Competition website
     \begin{center}
       \hcref{https://mccompetition.org/2022/mc\_description.html}
     \end{center}
     \begin{itemize}
     \item 100 public instances from Track 1 (unweighted model counting)
     \item 100 public instances from Track 2 (weighted model counting)
     \end{itemize}
   \item Removed duplicates
     \begin{itemize}
    \item  20 instances were duplicated in the two tracks
     \item Reduces set of benchmarks to 180
     \end{itemize}
   \item Ran D4 on 180 instances with time limit of 4000 seconds
     \begin{itemize}
     \item Completed for 124 instances
     \end{itemize}
   \item Generated POG representations from dDNNF graphs
     \begin{itemize}
     \item Simplifications via constant propagation
       \begin{itemize}
       \item Number of POG nodes around $0{.}90\times$ number of dDNNF nodes
       \end{itemize}
     \item Used heuristics to identify root node, and then put in topological ordering
     \item Measure complexity by number of defining clauses $D$
       \begin{itemize}
       \item Product or sum node with degree $k$ requires $k+1$ clauses
       \item $D$ also equals sum of number of edges and number of nodes
       \item For 124 instances, Min = 304, Median = 848,784, Mean = 93,107,937, Max = 2,761,457,765
       \end{itemize}
     \end{itemize}
   \item Ran program {\sc d2p} to generate CRAT file for maximum of 10,000 seconds
     \begin{itemize}
     \item Completed for 108 instances
     \item Exceeded clause limit for 1 instance
     \item Ran out of memory for 1 instance
     \item Timed out for 14 instances
     \end{itemize}
   \item For completed CRAT files, ran {\sc crat-check} CRAT checker
     \begin{itemize}
     \item Tests all conditions for correct CRAT
     \item Most time spent performing RUP checks for clause addition and deletion
     \item Completed successfully for all 108 instances
     \item Time generally faster than {\sc d2p}. (min = 0.005, max = 1.4, harmonic mean = 0.10)
     \end{itemize}
\end{enumerate}

@@

\begin{itemize}
\item Proof Clauses: Min 554, Median 1,719,245, Average 45,001,945, Max 1,719,245
\item Ratio to number of defining clauses: Min 1.54, Harmonic Mean 3.13, Max 6073
\end{itemize}

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 1.00]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000, 
                              x post scale=2.0, y post scale=2.0,
                              xmode=log,xmin=100,xmax=1000000000, 
                              xtick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, xticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
                              xlabel={Defining Clauses}, ylabel={Proof Clauses},
%                              title={D4 Defining Clause Generation}
            ]
    \addplot[mark=none, color=lightblue] coordinates{(100,100) (1e9,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,1000) (1e8,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,10000) (1e7,1e9)};
    \input{data-formatted/defining+total}
    \node[left] at (axis cs: 900,1200) {$1\times$};
    \node[left] at (axis cs: 900,12000) {$10\times$};
    \node[left] at (axis cs: 900,120000) {$100\times$};

@@

\begin{itemize}
\item Min 0.01, Average 1318.3, Max 13,144.8, Median 71.64
\item Ratio to time for D4: Min 0.50, Harmonic Mean 5.5, Max 457.7
\end{itemize}

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 1.00]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=100000,
                              x post scale=2.0, y post scale=2.2,
                              xmode=log,xmin=0.01,xmax=10000,
                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
                              ytick={0.01, 0.1,1.0,10,100,1000,10000,100000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000},{10,000},{100,000}},
                              xlabel={D4 runtime (seconds)}, ylabel={CRAT generation and checking runtime (seconds)},
%                              title={D4 Defining Clause Generation}
            ]
    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (10000,100000)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (1000,100000)};
    \input{data-formatted/time-d4-verify}
    \node[left] at (axis cs: 0.09,0.12) {$1\times$};
    \node[left] at (axis cs: 0.09,1.2) {$10\times$};
    \node[left] at (axis cs: 0.09,12.0) {$100\times$};

@@

Two optimizations were implemented within the CRAT framework:
\begin{description}
\item[Lemmas:] The justification proof must show that the conjunction
  of the input clauses implies the extension variable associated with
  the root node.  It does so by showing that this property holds for
  every node and for every path from the root that node, relative to
  the partial assignment given by that path.  This could potentially
  require enumerating exponential numbers of paths to some nodes,
  effectively converting the graph into a tree.  Instead, the proof
  can contain a {\em lemma} for each node having indegree greater than
  one, showing that the root node is implied by a set of clauses.
  Some of these are the original input clauses, while some are {\em
    synthetic clauses} representing input clauses that have been
  reduced by eliminating false literals.  Given such a proof, the
  lemma can be instantiated for each parent node.  Proofs with lemmas
  incur some overhead---the synthetic clauses are declared as POG
  product nodes, and additional clauses must be added to show that,
  under a given context, the synthetic clauses are implied by some
  input or containing synthetic clause.  On the other hand, they
  guarantee that no node in the graph is the root of more than one
  justication proof.
\item[Literal grouping:] The generated POG representation will
  typically have a number of product nodes forming the conjunction of
  tens, or even hundreds, of literals.  These literals represent ones
  that we either implied by unit propagation or were added when
  assigning the complement to the corresponding variable led to a
  conflict.  Justifying this product node requires showing that each
  of these literals must hold under the given context.  When this
  cannot be done by unit propagation, the program must invoke a SAT
  solver to prove that the complemented value yields a conflict.
  Rather than doing so for each literal individually, the prover can
  create a new product node consisting of the literals that have not
  yet been validated and then prove that falsifying the extension
  variable associated with the product node yields a conflict.  
\end{description}

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 1.00]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=1e9, 
      legend style={at={(0.40,0.98)}},
      legend cell align={left},
                              x post scale=2.0, y post scale=2.3,
                              xmode=log,xmin=1000,xmax=1e7, 
                              xtick={1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
                              ytick={1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
                              xlabel={Proof Clauses using Lemmas and Literal Grouping}, ylabel={Proof Clauses},
            ]
    \input{data-formatted/clauses-lemma-nolemma}
    \input{data-formatted/clauses-nosplit-split}
    \input{data-formatted/clauses-optimized-unoptimized}    
    \legend{
      \scriptsize \textsf{Without lemmas},
      \scriptsize \textsf{Without literal grouping},
      \scriptsize \textsf{Without either optimization},
    }
    \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
    \addplot[mark=none, color=lightblue] coordinates{(1000,10000) (1e8,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(1000,100000) (1e7,1e9)};

@@

          \end{axis}
\end{tikzpicture}
} % centering
\caption{Proof Clauses when Disable Options}
\label{fig:defining:unoptimized}

We conducted a set of experiments using the 80 benchmark instances for
which the total number of proof clauses when both optimizations are
performed was limited to $10^7$.  We then made three runs---with each
of the two optimizations disabled individually and in combination.
Figure \ref{fig:defining:unoptimized} illustrates the effect of the
two optimizations.  Each set of points shows the number of clauses
when both optimizations are performed for the $X$ value and the number
with one or both of these optimizations disabled as the $Y$ value.

The ratios between the number of clauses when either or both optimizations is disabled versus when both are enabled can be summarized as:
\begin{description}
\item[Without lemmas:] Min 1.00, Max 52.54, Harmonic mean 2.11
\item[Without literal grouping:] Min 0.84, Max 39.60, Harmonic mean 1.18
\item[Without either optimization:] Min 0.99, Max 52.54, Harmonic mean 2.95
\end{description}
Put another way: None of the 80 benchmarks require more than $10^7$
proof clauses when both optimizations are used.  Without lemmas, 19
rise above that threshold.  Without literal merging, 7 due.  Without
either, 25 rise above that threshold and 2 even exceed $10^8$ proof
clauses.

@@

>>>>

\section*{Methodology}
\begin{enumerate}
   \item Retrieved 200 public model counting benchmarks from 2022 Model Counting Competition website
     \begin{center}
       \hcref{https://mccompetition.org/2022/mc\_description.html}
     \end{center}
     \begin{itemize}
     \item 100 public instances from Track 1 (unweighted model counting)
     \item 100 public instances from Track 2 (weighted model counting)
     \end{itemize}
   \item Removed duplicates
     \begin{itemize}
    \item  20 instances were duplicated in the two tracks
     \item Reduces set of benchmarks to 180
     \end{itemize}
   \item Ran D4 on 180 instances with time limit of 4000 seconds
     \begin{itemize}
     \item Completed for 124 instances
     \end{itemize}
   \item Generated POG representations from dDNNF graphs
     \begin{itemize}
     \item Simplifications via constant propagation
       \begin{itemize}
       \item Number of POG nodes around $0{.}90\times$ number of dDNNF nodes
       \end{itemize}
     \item Used heuristics to identify root node, and then put in topological order with root node at end.
     \item Measure complexity by number of defining clauses $D$
       \begin{itemize}
       \item Product or sum node with degree $k$ requires $k+1$ clauses
       \item $D$ also equals sum of number of edges and number of nodes
       \item For 124 instances, Min = 304, Median = 848,784, Mean = 93,107,937, Max = 2,761,457,765
       \end{itemize}
     \end{itemize}
   \item Ran program {\sc d2p} to generate CRAT file for maximum of 10,000 seconds
     \begin{itemize}
     \item Completed for 108 instances
     \item Exceeded clause limit for 1 instance (Our programs represent a clause ID as a 32-bit signed value)
     \item Ran out of memory for 1 instance
     \item Timed out for 14 instances
     \end{itemize}
   \item For completed CRAT files, ran {\sc crat-check} CRAT checker
     \begin{itemize}
     \item Tests all conditions for correct CRAT
     \item Most time spent performing RUP checks for clause addition and deletion
     \item Completed successfully for all 108 instances
     \item Time generally faster than {\sc d2p}. (min = 0.005, max = 1.4, harmonic mean = 0.10)
     \end{itemize}
\end{enumerate}

@@

\begin{itemize}
\item Proof Clauses: Min 554, Median 1,719,245, Average 45,001,945, Max 770,382,773
\item Ratio to number of defining clauses: Min 1.54, Harmonic Mean 3.13, Max 6072.96
\end{itemize}

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 1.00]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000, 
                              x post scale=2.0, y post scale=2.0,
                              xmode=log,xmin=100,xmax=1000000000, 
                              xtick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, xticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
                              xlabel={Defining Clauses}, ylabel={Proof Clauses},
%                              title={D4 Defining Clause Generation}
            ]
    \addplot[mark=none, color=lightblue] coordinates{(100,100) (1e9,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,1000) (1e8,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,10000) (1e7,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,100000) (1e6,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,1000000) (1e5,1e9)};
    \input{data-formatted/defining+total}
    \node[left] at (axis cs: 900,1200) {$1\times$};
    \node[left] at (axis cs: 900,12000) {$10\times$};
    \node[left] at (axis cs: 900,120000) {$100\times$};
    \node[left] at (axis cs: 900,1200000) {$1{,}000\times$};
    \node[left] at (axis cs: 900,12000000) {$10{,}000\times$};

@@

\begin{itemize}
\item Min 0.01, Average 1318.3, Max 13,144.8, Median 71.64
\item Ratio to time for D4: Min 0.50, Harmonic Mean 5.50, Max 457.69
\end{itemize}

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 1.00]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=100000,
                              x post scale=2.0, y post scale=2.2,
                              xmode=log,xmin=0.01,xmax=10000,
                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
                              ytick={0.01, 0.1,1.0,10,100,1000,10000,100000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000},{10,000},{100,000}},
                              xlabel={D4 runtime (seconds)}, ylabel={CRAT generation and checking runtime (seconds)},
%                              title={D4 Defining Clause Generation}
            ]
    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (10000,100000)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (1000,100000)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (1000,1000000)};
    \input{data-formatted/time-d4-verify}
    \node[left] at (axis cs: 0.09,0.12) {$1\times$};
    \node[left] at (axis cs: 0.09,1.2) {$10\times$};
    \node[left] at (axis cs: 0.09,12.0) {$100\times$};
    \node[left] at (axis cs: 0.09,120.0) {$1000\times$};

@@

Two optimizations were implemented within the CRAT framework:
\begin{description}
\item[Lemmas:] Nodes with indegree greater than one were validated with separate lemmas, avoiding the possibly exponential expansion of the POG into a tree during the proof construction.

\item[Literal grouping:] For a product node with multiple literal
  arguments, those that cannot be validated by BCP are combined as
  arguments to a newly defined product node.  The extension variable
  associated with this node is validated, and the proof is then used
  to validate the literals.  This reduces the number of
  calls to the SAT solver to at most one per product node.
\end{description}

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 1.00]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=1e9, 
      legend style={at={(0.40,0.98)}},
      legend cell align={left},
                              x post scale=2.0, y post scale=2.3,
                              xmode=log,xmin=1000,xmax=1e7, 
                              xtick={1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
                              ytick={1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
                              xlabel={Proof Clauses using Lemmas and Literal Grouping}, ylabel={Proof Clauses},
            ]
    \input{data-formatted/clauses-lemma-nolemma}
    \input{data-formatted/clauses-nosplit-split}
    \legend{
      \scriptsize \textsf{Without lemmas},
      \scriptsize \textsf{Without literal grouping},
    }
    \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
    \addplot[mark=none, color=lightblue] coordinates{(1000,10000) (1e8,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(1000,100000) (1e7,1e9)};

@@

          \end{axis}
\end{tikzpicture}
} % centering
\caption{Proof Clauses when Single Optimization Disabled}
\label{fig:optimized:partially-optimized}
\end{figure}

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 1.00]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=1e9, 
      legend style={at={(0.40,0.98)}},
      legend cell align={left},
                              x post scale=2.0, y post scale=2.3,
                              xmode=log,xmin=1000,xmax=1e7, 
                              xtick={1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
                              ytick={1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
                              xlabel={Proof Clauses using Lemmas and Literal Grouping}, ylabel={Proof Clauses},
            ]
    \input{data-formatted/clauses-optimized-unoptimized}    
    \legend{
      \scriptsize \textsf{Without either optimization},
    }
    \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
    \addplot[mark=none, color=lightblue] coordinates{(1000,10000) (1e8,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(1000,100000) (1e7,1e9)};

          \end{axis}
\end{tikzpicture}
} % centering
\caption{Proof Clauses Both Optimizations are Disabled}
\label{fig:optimized:unoptimized}
\end{figure}

We conducted a set of experiments using the 80 benchmark instances for
which the total number of proof clauses when both optimizations are
performed was limited to $10^7$.  We then made three additional runs
of the CRAT generator and checker, with each of the two optimizations
disabled individually and in combination.  Figures
\ref{fig:optimized:partially-optimized}--\ref{fig:optimized:unoptimized} illustrates the effect of the two
optimizations.  Each set of points shows the number of clauses when
both optimizations are performed for the $X$ value and the number with
one or both of these optimizations disabled as the $Y$ value.

The ratios between the number of clauses when either or both optimizations is disabled versus when both are enabled can be summarized as:
\begin{description}
\item[Without lemmas:] Min 1.00, Max 52.54, Harmonic mean 2.11
\item[Without literal grouping:] Min 0.84, Max 39.60, Harmonic mean 1.18
\item[Without either optimization:] Min 0.99, Max 52.54, Harmonic mean 2.95
\end{description}

Here are other ways to summarize the impact on proof length:
\begin{itemize}
\item None of the 80 benchmarks require more than $10^7$
proof clauses when both optimizations are used.  Without lemmas, 19
rise above that threshold.  Without literal merging, 7 due.  Without
either, 25 rise above the threshold.  Two of the benchmarks require
more than $10^8$ proof clauses without lemmas.
\item Of the 80 benchmark problems evaluated, and comparing with both
  optimizations disabled, 55 had the number of proof clauses reduced
  by at least 10\% by using lemmas, 19 by using literal merging, and
  69 by enabling both optimizations.
\end{itemize}

The time performance is less dramatic, but still significant.
Comparing the sum of CRAT generation and checking times for each of
the four combinations of options, we get the following times relative
to the one with both optimizations enabled:
\begin{description}
\item[Without lemmas:] Min 0.85, Max 37.98, Harmonic mean 1.80
\item[Without literal grouping:] Min 0.83, Max 3.92, Harmonic mean 1.12
\item[Without either optimization:] Min 0.91, Max 41.97, Harmonic mean 2.19
\end{description}

Here is another way to summarize the impact on the combined time to generate and validate the CRAT file:
\begin{itemize}
\item Of the 80 benchmark problems evaluated, and comparing with both
  optimizations disabled, 50 had the time reduced by at least 10\% by
  using lemmas, 27 by using literal merging, and 63 by enabling both
  optimizations.
\end{itemize}

@@

\begin{description}
\item[Lemmas:] The justification proof must show that the conjunction
  of the input clauses implies the extension variable associated with
  the root node.  It does so by showing that this property holds for
  every node and for every path from the root that node, relative to
  the partial assignment given by that path.  This could potentially
  require enumerating exponential numbers of paths to some nodes,
  effectively converting the graph into a tree.  Instead, the proof
  can contain a {\em lemma} for each node having indegree greater than
  one, showing that the root node is implied by a set of clauses.
  Some of these are the original input clauses, while some are {\em
    synthetic clauses} representing input clauses that have been
  reduced by eliminating false literals.  Given such a proof, the
  lemma can be instantiated for each parent node.  Proofs with lemmas
  incur some overhead---the synthetic clauses are declared as POG
  product nodes, and additional clauses must be added to show that,
  under a given context, the synthetic clauses are implied by some
  input or containing synthetic clause.  On the other hand, they
  guarantee that no node in the graph is the root of more than one
  justication proof.
\item[Literal grouping:] The generated POG representation will
  typically have a number of product nodes forming the conjunction of
  tens, or even hundreds, of literals.  These literals represent ones
  that we either implied by unit propagation or were added when
  assigning the complement to the corresponding variable led to a
  conflict.  Justifying this product node requires showing that each
  of these literals must hold under the given context.  When this
  cannot be done by unit propagation, the program must invoke a SAT
  solver to prove that the complemented value yields a conflict.
  Rather than doing so for each literal individually, the prover can
  create a new product node consisting of the literals that have not
  yet been validated and then prove that falsifying the extension
  variable associated with the product node yields a conflict.  
\end{description}
